### 2020-11-12 latex/sections/methods.tex
latexing

In this chapter, we describe our implementation of the game Breakthrough and the methods for training a neural network to play the game Breakthrough, described in Chapter \ref{cha:background}. The training process uses MCTS described in Section \ref{sec:mcts} to guide its training. We outline the architecture used in the neural network.

@@

In this section, we will take a brief look at the implementation of the game Breakthrough, how we model its state representation and the pseudocode for training.

\subsection{Breakthrough}

For a neural network to be able to use input, that input needs to be numeric. We model the board in gamethrough with a single $NxM$ array with three values $\{-1,0,1\}$. Where each $x,y$ index on the board represents what value is on the board at the time, $-1$ represents a white piece, $0$ an empty square, and $1$ a black piece.

An action vector is coupled with each state, the action vector is a $3$-dimensional matrix where the lengths of the dimensions are $N * M * 6$ and each index of the matrix represents an action moving from cell $x,y$ moving to the direction $z$. The directions are as follows $0$ represents moving upwards and to the left diagonally on the board, $1$ upwards, and $2$ upwards and to the right diagonally, $4$ downwards and to the left diagonally, $5$ downwards, and lastly $6$ downwards and to the right diagonally.

@@

The neural network architecture we opted to use was a single convolutional layer with a ReLU activation function, followed by 5 residual layers each containing two layers of convolution, batch normalization, and a ReLu activation. Lastly, a split policy/value head, where the output of the last residual layer is split into two different outputs. The policy head are two layers: first a final convolutional layer, and then a fully connected layer with a \textit{log softmax} layer. The value head consists of three layers: first a convolutional layer, then a fully connected layer, and lastly a fully connected layer with a \textit{tanh} activation function. The Figure \ref{fig:nnarch}. depicts the architecture.

@@

It is important to understand why we use convolutional layers when dealing with board games as convolution is often associated with an image. This is because a convolutional layer is focused on merging multiple input parameters to a single neuron, making it an input parameter for the next layer representing the locality around the center point of the original input. Playing the game of Breakthrough, a piece is only able to capture pieces in its immediate vicinity. This lead to the selection of a convolutional kernel with a size of $3x3$ and a stride of $1$.

The knowledge of each piece moves isn't enough to represent the whole game state within the neural network. This is why we use multiple residual layers with convolution. These residual layers stack convolutions on each other making the final convolution represent the locality of all the other localities, allowing the neural network to have a representation of the whole game state in its parameters.

@@

The residual layers as a concept remember the output of the previous layer and add the results of the residual layer and the previous layer together. On a higher level, this leads to the internal representation of the neural network to maintain the whole game boards in its representation, s.t. two areas that are far away from each other maintain the same level of locality as two that are close.

@@

This subsection focuses on the implementation of the various functions required to train the neural network to play breakthrough. The description is mainly through pseudo code, and the code is awailable on \href{github.com/sigurdurhelga/msc-cheis}{Github} for examination.

@@

\begin{algorithm}[t]
\caption{Neural network selfplay pseudocode}
\label{alg:selfplay}
\begin{algorithmic}[1]
\STATE \textbf{neural network} nn1 = randomizedInitialNN()
\STATE \textbf{neural network} nn2 = randomizedInitialNN()
\WHILE{\TRUE}
    \STATE{dataset = \textbf{generate\_dataset}(nn1)}
    \STATE{nn1.train\_on\_examples(dataset)}
    \STATE{win\_rate = \textbf{compete}(nn1,nn2)}
    \IF{win\_rate > $0.5$}
        \STATE{nn2 = nn1.copy()}
    \ELSE
        \STATE{nn1 = nn2.copy()}
    \ENDIF
    \STATE \textbf{save}(nn1)
\ENDWHILE
\end{algorithmic}
\end{algorithm}

To train the neural network to play Breakthrough we initialize two neural networks nn1 and nn2 with random weights. Then we let nn1 play against itself using MCTS with PUCT to select moves. We collect data from this self-play to train on. The data collected is $(s, \pi, \tau, p, v)$, $s$ is the state, $\pi$ is the action probability vector provided by MCTS with PUCT, $\tau$ the end reward for the episode $1$ if white wins, $-1$ if black wins. $p$ is the action probability vector predicted by the neural network, and $v$ is the predicted reward of the game by the neural network.

The pseudocode can be seen in Algorithm~\ref{alg:selfplay}.

@@

\begin{algorithm}[t]
\caption{Neural network compete pseudocode}
\label{alg:competition}
\begin{algorithmic}[1]
\STATE{\textbf{Input:} neural network nn1, neural network nn2}
\STATE{\textbf{Output:} win rate for white player}
\STATE{white\_wins $= 0$}
\STATE{game = \textbf{initial\_breakthrough}()}
\FOR{1 \TO 100}
    \WHILE{\textbf{not} game.is\_termina()}
        \STATE{nn1.make\_move()}
        \STATE{nn2.make\_move()}
    \ENDWHILE
    \IF{game.white\_wins()}
        \STATE{white\_wins++}
    \ENDIF
\ENDFOR
\STATE{\textbf{return} white\_wins/100}
\end{algorithmic}
\end{algorithm}

The compete function differs from the self-play function in such a way that we select the moves by a single pass through the neural network thereby only evaluating the neural networks ability to predict best actions. The pseudocode is available in Algorithm~\ref{alg:competition}.

@@

Once the neural network has learned to play the game of Breakthrough, we examine its internal state w.r.t HLC's that we understand. The first examined HLC is numbers advantage, that is, the number of pieces that a player has over his opponent. The higher-level idea to a human player would be that they are in a better position since they have more pieces.

To examine the higher-level component we take a look at the internal state of the neural network itself. To do this we take a state, we run the state through the neural network, and while the state is propagating through the network we select a layer to split the network, we call that layer $l_{split}$. At that point in time the state is a mutated vector $l_{split}(l_{split-1}(...l_0(state)))$ of the initial state. We save that vector as a point in the N-dimensional space that it represents, and once we've gathered enough points in that N-dimensional space we're able to train a linear classifier using the HLC as a label. We train a Stochastic Gradient Descent classifier, from the SKLearn library, to construct a hyperplane that splits the space into two binary states, one that contains the HLC and another that doesn't.

@@

An example of how this would look with only two dimensions is shown in Figure \ref{fig:scattersplit} Once we've successfully trained this linear classifier we can run another state $s$ through the neural network. Once the prediction has finished, we view the selected action $a$ of $s$ from the policy vector $p$ from the neural network. We view the loss of that selected action $a$. Then we apply the backpropagation algorithm up to that same layer $l_{split}$ and evaluate the gradient there. If that gradient moves in the direction of the HLC we say that the state includes the HLC, and doesn't if it moves away from the HLC.

@@

We show an example in Figure \ref{fig:scattersplitarrow}. where the arrow in the image represents the direction the gradient is moving, if it moves toward the HLC the state $s$ includes the HLC otherwise it does not. Using this method we evaluate the internal representation of the state within the neural network and can see whether it recognizes the HLC.

>>>>

In this chapter, we describe our implementation of the game Breakthrough and the methods for training a neural network to play the game. The training process uses MCTS, as described in Section \ref{sec:mcts}, to guide its training. We outline the architecture used in the neural network.

@@

In this section, we will take a look at the implementation of the Breakthrough neural network, how we model its state representation and the pseudo-code for training.

\subsection{Description of the input and output}

For a neural network to be able to use input, that input needs to be numeric. We model the board with a single $NxMx3$ array with two values $\{0,1\}$, where the first two indices on the $z$ axis represent the players, and the last layer on the $z$ axis represents which players turn it is. The first $Z$ index represents whites pawns, that is, each $x,y$ position on the board where white has a pawn that cell $x,y,0$ has the value $1$ otherwise it has $0$, and similarily on the second layer for the black player. The last $z$ layer, contains all $1$'s if it is the white players turn otherwise $0$.

One of the outputs of the nerual network is the policy vector, the policy vector is a $3$-dimensional matrix where the lengths of the dimensions are $N * M * 6$ and each index of the matrix represents an action moving from cell $x,y$ moving to the direction $z$. The directions are as follows $0$ represents moving upwards and to the left diagonally on the board, $1$ upwards, and $2$ upwards and to the right diagonally, $4$ downwards and to the left diagonally, $5$ downwards, and lastly $6$ downwards and to the right diagonally. The cell values indicate the probability distribution of the neural network selecting that move. Importantly, before we select a move from the neural networks prediction we zero out the illegal moves, and renormalize the array such that it sums to $1$.

@@

The neural network architecture we opted to use was a single convolutional layer with a ReLU activation function, followed by 5 residual layers each containing two layers of convolution, batch normalization, and a ReLu activation. Lastly, a split policy/value head, where the output of the last residual layer is split into two different outputs. The policy head are two layers: first a final convolutional layer, and then a fully connected layer with a \textit{log softmax} layer. The value head consists of three layers: first a convolutional layer, then a fully connected layer, and lastly a fully connected layer with a \textit{tanh} activation function. Figure~\ref{fig:nnarch}. depicts the architecture. Using this architecture we end up with $1737986$ trainable parameters.

@@

It is important to understand why we use convolutional layers when dealing with board games as convolution is more typically associated with an image. This is because a convolutional layer is focused on merging multiple input parameters to a single neuron, making it an input parameter for the next layer representing the locality around the center point of the original input. Playing the game of Breakthrough, a piece is only able to capture pieces in its immediate vicinity. This lead to the selection of a $3$-d convolutional kernel with a size of $3x3$ and a stride of $1$.

We require a global view of the board. This is why we use multiple residual layers with convolution. These residual layers stack convolutions on each other making the final convolution represent the locality of all the other localities, allowing the neural network to have a representation of the whole game state in its parameters.

@@

The residual layers as a concept remember the output of the previous layer and add together the results of the residual layer and the previous layer. On a higher level, this leads to the internal representation of the neural network to maintain the whole game boards in its representation, s.t. two areas that are far away from each other maintain the same level of locality as two that are close.

@@

This subsection focuses on the implementation of the various functions required to train the neural network to play Breakthrough. The description is mainly through pseudo-code, and the code is available on GitHub for examination~\cite{siggi:github}.

@@

\begin{algorithm}[t]
    \caption{Neural network self-play pseudo-code}
    \label{alg:selfplay}
    \begin{algorithmic}[1]
        \STATE \textbf{neural network} nn1 = randomizedInitialNN()
        \STATE \textbf{neural network} nn2 = randomizedInitialNN()
        \WHILE{\TRUE}
        \STATE{dataset = \textbf{generate\_dataset}(nn1)}
        \STATE{nn1.train\_on\_examples(dataset)}
        \STATE{win\_rate = \textbf{compete}(nn1,nn2)}
        \IF{win\_rate > $0.5$}
        \STATE{nn2 = nn1.copy()}
        \ELSE
        \STATE{nn1 = nn2.copy()}
        \ENDIF
        \STATE \textbf{save}(nn1)
        \ENDWHILE
    \end{algorithmic}
\end{algorithm}

To train the neural network to play Breakthrough we initialize two neural networks nn1 and nn2 with random weights. Then we let nn1 play against itself using MCTS with PUCT to select moves. We collect data from this self-play to train on. The data collected is $(s, \pi, \tau, p, v)$, where $s$ is the state, $\pi$ is the action policy vector provided by MCTS with PUCT, $\tau$ the end reward for the episode $1$ if white wins, $-1$ if black wins. $p$ is the policy vector predicted by the neural network, and $v$ is the predicted reward of the game by the neural network.

The pseudo-code is shown as Algorithm~\ref{alg:selfplay}.

@@

\begin{algorithm}[t]
    \caption{Neural network compete pseudo-code}
    \label{alg:competition}
    \begin{algorithmic}[1]
        \STATE{\textbf{Input:} neural network nn1, neural network nn2}
        \STATE{\textbf{Output:} win rate for white player}
        \STATE{white\_wins $= 0$}
        \STATE{game = \textbf{initial\_breakthrough}()}
        \FOR{1 \TO 100}
        \WHILE{\textbf{not} game.is\_termina()}
        \STATE{nn1.make\_move()}
        \STATE{nn2.make\_move()}
        \ENDWHILE
        \IF{game.white\_wins()}
        \STATE{white\_wins++}
        \ENDIF
        \ENDFOR
        \STATE{\textbf{return} white\_wins/100}
    \end{algorithmic}
\end{algorithm}

The compete function differs from the self-play function in such a way that we select the moves by a single pass through the neural network thereby only evaluating the neural networks ability to predict best actions. The pseudo-code shown as Algorithm~\ref{alg:competition}.

@@

Once the neural network has learned to play the game of Breakthrough, we examine its internal state w.r.t HLC's that we understand. The first examined HLC is material advantage, that is, the number of pieces that a player has over the opponent. The higher-level idea to human players would be that they are in a better position since they have more pieces.

To examine the higher-level component we take a look at the internal state of the neural network itself. To do this, we take a state, and run it through the neural network, and while the state is propagating through the network we select a layer to split the network, we call that layer $l_{split}$. At that point in time the state is a mutated vector $l_{split}(l_{split-1}(...l_0(state)))$ of the initial state. We save that vector as a point in the N-dimensional space that it represents, and once we've gathered enough points in that N-dimensional space we're able to train a linear classifier using the HLC as a label. We train a Stochastic Gradient Descent classifier, from the SKLearn library, to construct a hyper-plane that splits the space into two binary states, one that contains the HLC and another that doesn't.

@@

An example of how this would look with only two dimensions is shown in Figure \ref{fig:scattersplit}. Once we've successfully trained this linear classifier we can run another state $s$ through the neural network. Once the prediction has finished, we view the selected action $a$ of $s$ from the policy vector $p$ from the neural network. We view the loss of that selected action $a$. We then apply the backpropagation algorithm up to that same layer $l_{split}$ and evaluate the gradient there. If that gradient moves in the direction of the HLC we say that the state includes the HLC, and that it doesn't if the gradient moves away from the HLC.

@@

We show an example in Figure \ref{fig:scattersplitarrow}, where the arrow in the image represents the direction the gradient is moving, if it moves toward the HLC the state $s$ includes the HLC otherwise it does not. Using this method we evaluate the internal representation of the state within the neural network and can see whether it recognizes the HLC.

\section{Summary}

In this chapter we described the architecture of the nerual network, including how we model Breakthrough to be able to learn how to play the game. We described the processes we use to learn completely by self play. In the following chapter we will evaluate the neural network.

### 2020-11-12 latex/sections/results.tex
latexing

\ifdraft{In this section you discuss any issues that came up while developing
the system.  If you found something particularly interesting,
difficult, or an important learning experience, put it here.  This is
also a good place to put additional figures and data.}
hello world

>>>>

\label{cha:results}

In this chapter, we first evaluate the neural network against various agents to validate that it has learned how to play the game. Secondly, we describe each concept we intend to examine the neural network against. Thirdly, we take a look at the changes in the emphasis of the neural network during training. The main point of interest there being whether the neural network notices simple HLCs early then stops considering them as the network improves.

\section{Evolution of the Neural Networks Win Rate}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsrandom.png}
        \end{center}
        \caption{Winrate vs random agent}
        \label{fig:winratevsrandom}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsmcts.png}
        \end{center}
        \caption{Winrate vs MCTS agent}
        \label{fig:winratevsmcts}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsneuralnetwork.png}
        \end{center}
        \caption{Winrate vs neural network trained for 150 generations}
        \label{fig:winratevsneuralnetwork}
    \end{small}
\end{figure}

Examining the neural network is only interesting if the neural network can effectively play the game. We first examine the history neural network playing against an agent that only takes random moves, on every $10$ generations. The win rate over generations can be seen in Figure \ref{fig:winratevsrandom}. It is clear that very early on in the training process, almost immediately after generation $10$ our agent performs nearly perfectly against the random agent. This implies that the agent does indeed learn some strategy in the game. The next agent we tried the neural network against, was an agent that does regular Monte-Carlo tree search on the game space with $100$ iterations of rollout for each move. A graph showing the win rate over generations is depicted in Figure \ref{fig:winratevsmcts}. Again, our agent quickly learns some strategy and can win often, although doesn't achieve a perfect win rate after $300$ generations. The last agent we tested against was the neural network itself, although only trained to $150$ generations. The graph in Figure \ref{fig:winratevsneuralnetwork} shows win rate. As expected the agent performs poorly until it has trained for more than $150$ generations.

These results imply that our agent certainly understands how to play the game of breakthrough to a certain level. However, an intermediate level human player would most often be able to win the agent, based on the author's experience playing against it. Note, that the main focus of this was not to create an expert level agent. Rather to, gauge into the concepts the agent learns as described in the next section.

\section{Evaluating the improvements of the Neural Network over generations}

To test which HLC the neural network places its emphasis on during training we trained a neural network for $300$ generations, taking snapshots of the network every $10$ generations. We then had the neural network play against itself for $100$ games, collecting the states it encountered during play. These states were then examined by a concept activation vector representing these HLCs.

We test four concepts \textit{Material Advantage}, \textit{Aggressiveness}, \textit{Unity}, and \textit{Lorentz-Horey}.

\subsection{Description of concepts}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_number_adv}
        \end{center}
    \end{small}
    \caption{Distribution of piece amount difference}
    \label{fig:distnumberadv}
\end{figure}

The first concept we examine is Material Advantage which conveys the idea of having more pawns than your opponent. When faced with a board it is difficult to argue for whether having a single pawn up on your opponent constitues as Material Advantage or three. In order to an appropriate value for this case we used the neural network that had been trained for $300$ generations to generate $20000$ unique states. The distribution of these states' difference of pawns can be seen in Figure \ref{fig:distnumberadv}. From examining the distribution and examining samples we selected the breakpoint of $\geq 2$, meaning that if you have $2$ pieces on your opponent you are in a state that has this concept. From the $20000$ states only $5.5\%$ of states have this concept.

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_aggressiveness}
        \end{center}
        \caption{Distribution of difference of furthest pawns}
        \label{fig:distaggressiveness}
    \end{small}
\end{figure}

The next concept is Aggressiveness, which describes how much closer your furthest piece is to the opponents edge than your opponents furthest piece to your edge. This is then the minimum amount of how many moves it will take you to win the game. A distribution of these values can be seen in Figure \ref{fig:distaggressiveness}. We examined the distribution, and sampled states from various breaking points and decided on the value of $\geq 1$. That is if the value of a state's furthest piece difference is greater than or equal to $1$ that state has the concept of Aggressiveness. From the dataset $18.3\%$ of states have this concept.

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_unity}
        \end{center}
        \caption{Difference of distance from center}
        \label{fig:distunity}
    \end{small}
\end{figure}

Our third concept is Unity. This concept represents the absolute average distance of your pawns from the center row of your pawns, and is calculated as follows: the row of your furthest pawn from the starting row $r_{far}$, the row of your nearest pawn from the starting row $r_{near}$. Finding the middle row is then $\frac{r_{far} + r_{near}}{2}$, we then take the absolute of the average distance from all of the players pawns to that row. The distribution of the average distance values is shown in Figure \ref{fig:distunity}. From sampling states from several points we decided on a breakpoint of $\le 0.25$ for identifying states as states where the Unity concept is precent. The value of $0.25$ generally allows your states to have two rows that have the majority of the pawns and one or two pawns one row away from the group. From the dataset of 20,000 states $20\%$ of states have this concept.

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_lorentz}
        \end{center}
        \caption{Difference of Lorentz-Horey Score}
        \label{fig:distlorentz}
    \end{small}
\end{figure}

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        5  & 15 & 5  & 5  & 15 & 5  \\\hline
        2  & 3  & 3  & 3  & 3  & 2  \\\hline
        4  & 6  & 6  & 6  & 6  & 4  \\\hline
        7  & 10 & 10 & 10 & 10 & 7  \\\hline
        11 & 15 & 15 & 15 & 15 & 11 \\\hline
        21 & 21 & 21 & 21 & 21 & 21 \\\hline
    \end{tabular}
    \caption{Lorentz-Horey cell values, from white players point of view}
    \label{table:lorentzcell}
\end{table}

The last concept we examined was the Lorentz-Horey. This concept is a popular heuristic in Breakthrough defined in the paper by Lorentz \& Horey \cite{lorentz:heuristic}. For this heuristic, we give each cell on the board a point value. For a given player the heuristic is then the sum of the cell values where they have pawns. The cell values are shown in Table \ref{table:lorentzcell}. The matrix shown is flipped from the black player's point of view. The values are selected in such a way that as your pawn approach the opponent's side their value increases, having pieces on the edge isn't optimal, as such a pawn will only be able to capture in one direction and can not escape capture as easily. To select the breakpoint for Lorentz-Horey heuristic, we generate a distribution of the difference in Lorentz-Horey Score over our dataset. The resulting distribution is shown in Figure \ref{fig:distlorentz}. Our examination lead us to select the value of $5$, as a breakpoint indicating that a state has the concept of Lorentz-Horey. Out of our dataset of 20,000 states $29.4\%$ of states have this concept.

\subsection{Material Advantage}

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/number_pawns_trend}
    \caption{Percentage of selected states containing the HLC Material Advantage}
    \label{fig:numberadvantage}
\end{figure}

The Figure \ref{fig:numberadvantage}. shows that throughout training the Material Advantage HLC is only ever a slight factor in the selection of states and we can say that the neural network doesn't consider number advantage in its selection process.

\subsection{Agressiveness}

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/most_advanced_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:aggressiveness}
\end{figure}

From Figure \ref{fig:aggressiveness}. we can see that the aggressiveness HLC is a growing factor over as the neural network is trained. As a piece can always move forward, if your opponent does not capture that piece, this concept relates to how many moves it will take for you to win. Generally, when your opponent is not playing well this concept does well, and quickly becomes a bad heuristic in classical search algorithms.

\subsection{Unity}

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/unity_trend.png}
    \caption{Percentage of selected states containing the HLC unity}
    \label{fig:unity}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/lorentz_horey_trend.png}
        \end{center}
        \caption{Percentage of selected states containing the HLC Lorentz-Horey}
        \label{fig:lorentzheuristic}
    \end{small}
\end{figure}

The literature of Breakthrough implies that being patient and waiting until your opponent makes a mistake is generally the preferred strategy for winning.\cite{lorentz:heuristic} Examining Figure \ref{fig:lorentzheuristic}, we can see that the trend of using this HLC increases over time. As this heuristic is the one most often cited as a successful strategy in Breakthrough, it is interesting to see that our Breakthrough agent is gradually learning a similar concept.

\section{Summary}

In this section we discussed some concepts that we believed would be representative of concepts that a nerual network would learn over time playing Breakthrough. An important note is that there are an infinite amount of concepts, and there could exist some that the neural network uses extensively that we overlooked. We examined the neural network against these concepts. In the following chapter we will discuss the next steps for this research.
